Comparison of FDR methods on GWAS with SummarizedBenchmark

### 2017-12-14 datasets/GWAS/GWAS-SummarizedBenchmark.Rmd
First success at efficient LD pruning

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
# project directory & data/results folders
#setwd("/n/irizarryfs01_backed_up/kkorthauer/FDR")
datdir <- "/n/irizarryfs01_backed_up/kkorthauer/FDR/DATA/GWAS/"
resdir <- "/n/irizarryfs01_backed_up/kkorthauer/FDR/RESULTS/GWAS/"

# wrangling & plotting tools
library(data.table)
library(readxl)
library(dplyr)
library(ggplot2)
library(magrittr)
library(R.utils)
library(cowplot)

@@

```{bash, GWAS1-download, eval=FALSE}
cd /n/irizarryfs01_backed_up/kkorthauer/FDR/DATA/GWAS/
curl -O "http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip"
```

@@

```{bash, GWAS1-unzip, eval=FALSE}
unzip BMI.SNPadjSMK.zip
rm BMI.SNPadjSMK.zip
rm BMI.SNPadjSMK.Women.EuropeanOnly.txt
rm BMI.SNPadjSMK.Women.AllAncestry.txt
rm BMI.SNPadjSMK.Men.EuropeanOnly.txt
rm BMI.SNPadjSMK.Men.AllAncestry.txt
rm BMI.SNPadjSMK.CombinedSexes.AllAncestry.txt
```

@@

@@

# upset plot 
plotFDRMethodsOverlap(sb, 
                   alpha=0.05, nsets=ncol(sb),
                   order.by="freq", decreasing=TRUE,
                   supplementary=FALSE)
```

@@

# upset plot 
plotFDRMethodsOverlap(sb, 
                   alpha=0.05, nsets=ncol(sb),
                   order.by="freq", decreasing=TRUE,
                   supplementary=FALSE)
```

@@

```{r, GWAS1-cleanup}
rm(bmi)
rm(bd)
rm(sb)
rm(pf)
```

>>>>

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
# project directory & data/results folders - the datdir needs to match where
#  files are downloaded with the first few bash chunks
#setwd("/n/irizarryfs01_backed_up/kkorthauer/FDR")
datdir <- "DATA/GWAS/"
resdir <- "RESULTS/GWAS/"

# wrangling & plotting tools
library(data.table)
library(readxl)
library(readr)
library(dplyr)
library(ggplot2)
library(magrittr)
library(R.utils)
library(cowplot)

@@

```{bash, GWAS1-download, eval=FALSE}
curl -o DATA/GWAS "http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip"
```

@@

```{bash, GWAS1-unzip, eval=FALSE}
unzip DATA/GWAS/BMI.SNPadjSMK.zip
rm DATA/GWAS/BMI.SNPadjSMK.zip
rm DATA/GWAS/BMI.SNPadjSMK.Women.EuropeanOnly.txt
rm DATA/GWAS/BMI.SNPadjSMK.Women.AllAncestry.txt
rm DATA/GWAS/BMI.SNPadjSMK.Men.EuropeanOnly.txt
rm DATA/GWAS/BMI.SNPadjSMK.Men.AllAncestry.txt
rm DATA/GWAS/BMI.SNPadjSMK.CombinedSexes.AllAncestry.txt
```

We also download the 1000 genomes CEU reference population data, which will be used to 
determine which SNPs are LD buddies for the purposes of obtaining an independent set of SNPs.

```{bash, GWAS1-refdata, eval=FALSE}
# download CEU 1000 genomes SNP reference data 
wget 'http://neurogenetics.qimrberghofer.edu.au/iSECA/1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01.zip' -P DATA/GWAS/

# unzip reference data
unzip DATA/GWAS/1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01.zip -d DATA/GWAS
```

@@

#### Prune SNPs that are in Linkage Disequilibrium

Since nearby SNPs in linkage disequilibrium may not constitute separate discoveries, we carry out 
a pruning step. This commonly done by examining correlation of genotypes among nearby markers, but
since we don't have the genotype data, we'll use the linkage disequilibrium results of common SNPs and 
previously observed LD blocks. Specifically, we'll use the phase 3 data from 1000 genomes, restricted
to the CEU population (since this most closely matches our population of European ancestry). This 
reference data was downloaded in a previous code chunk. Here we read in the list of more than
9 million SNPs with data in 1000 genomes and create a list of SNPs that overlap our set. We'll write 
this list to a file that matches PLINK format.

```{r, GWAS1-snplist, eval=TRUE}
# load reference data 
onekg <- fread(paste0(datdir, "1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01.bim"))$V2

# construct input list of SNPs that overlap - 97.5% overlap
bmi$BP <- as.numeric(unlist(lapply(strsplit(bmi$markername, ":"), function(x) x[[2]])))
write_delim(bmi %>% dplyr::rename(CHR=chromosome, SNP=rs_id, A1=allele_1, A2=allele_2) %>%
              mutate(P=0.5, F_A=Freq_Allele1_HapMapCEU, F_U=F_A, CHISQ=1, OR=1) %>% 
              select(CHR, SNP, BP, A1, F_A, F_U, A2, CHISQ, P, OR) %>%
              filter(SNP %in% onekg), 
            path=paste0(datdir, "rslist.assoc"), delim="\t")

ol <- sum(bmi$rs_id %in% onekg)/nrow(bmi)
rm(onekg)
```

It turns out that `r signif(ol*100, 3)` percent of SNPs in our set are included in the 1000 genomes 
reference data. Next we'll use the PLINK software to read in our list of SNPs and the 1000 genomes
linkage disequilibrium data to 'clump' the SNPs into independent sets of SNPs that have LD correlation (r squared)
less than 0.2 with all neighbors within 250 kilobases of distance. We write the results to a file called `plink_clump.clumped`. We have to jump out of R here since PLINK is a command line tool.

```{bash, clump, eval=FALSE}
# use PLINK clumping tool to obtain a set of independent SNPs by LD
module load plink

plink \
--bfile DATA/GWAS/1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01 \
--clump DATA/GWAS/rslist.assoc \
--clump-field P \
--clump-p1 0.9999 \
--clump-p2 0.9999 \
--clump-r2 0.2 \
--clump-kb 250 \
--out DATA/GWAS/plink_clump

```

Next, we'll read in the clumped results and subset our data by our list of independent SNPs.

```{r, GWAS1-prune}
clump <- fread(paste0(datdir, "plink_clump.clumped"))

# subset to include only the SNPs in the ld.list
bmi <- bmi[bmi$rs_id %in% clump$SNP,]

rm(clump)
```

After pruning, we are left with for `r nrow(bmi)` approximately independent SNPs.

@@

# upset plot 
plotFDRMethodsOverlap(sb, 
                      alpha=0.05, nsets=ncol(sb),
                      order.by="freq", decreasing=TRUE,
                      supplementary=FALSE)

# properties of method-exclusive discoveries
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf" )
mcols(sb)$ind_covariate <- mcols(sb)$ind_covar_N
plotCovariateBoxplots(sb, alpha=0.1, nsets=6, methods=methods, trans=log2, maxNum=5)

```

@@

# upset plot 
plotFDRMethodsOverlap(sb, 
alpha=0.05, nsets=ncol(sb),
order.by="freq", decreasing=TRUE,
supplementary=FALSE)

# properties of method-exclusive discoveries
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf" )
mcols(sb)$ind_covariate <- abs(mcols(sb)$ind_covar_AF-0.5)
plotCovariateBoxplots(sb, alpha=0.1, nsets=6, methods=methods, trans=log2, maxNum=5)

```

@@

```{r, GWAS1-cleanup}
rm(bmi)
rm(bd)
rm(sb)
```
